---
title: lecture 4 - multiple testing correction - with code
author: "Haky Im"
date: "March 31, 2025"
description: "Lecture 4"
categories:
  - bios25328
  - notebook
freeze: true
---

Find the lecture notes [here](https://www.icloud.com/keynote/07fVVMHEvj8f2aG5zyyFfXisA#L4-Multiple-Testing-Correction-2025).

```{r setup, include=FALSE}
# Ensure necessary packages are loaded silently if needed later
# library(qvalue) 
# library(knitr)
# Set seed for reproducibility if desired globally
set.seed(20210108) 
```

# Learning Objectives

- Build intuition about p-values when multiple testing is performed via simulations.
- Recognize the need for multiple testing correction.
- Present methods to correct for multiple testing:
  - Bonferroni correction
  - FDR (false discovery rate)

# Why Do We Need Multiple Testing Correction?

When performing many statistical tests simultaneously, the probability of finding a "significant" result just by chance (a Type I error or false positive) increases substantially. This is famously illustrated by the XKCD comic on significance (link to XKCD #882). We need methods to control the overall error rate across all tests.

# What Do P-values Look Like Under Null and Alternative Hypotheses?

## Simulate Data: Null vs. Alternative

We'll start by simulating data under two scenarios:

- Null Hypothesis (Ynull): The outcome variable is independent of the predictor X.
- Alternative Hypothesis (Yalt): The outcome variable depends on the predictor X via the relationship Yalt = X * beta + epsilon.

## Simulation Parameters

First, we will define some parameters for the simulation:

```{r}
# Parameters
nsamp = 100     # Number of samples
beta = 2        # Effect size for the alternative hypothesis
h2 = 0.1        # Proportion of variance in Yalt explained by X (related to signal strength)
sig2X = h2      # Variance of X (scaled for simplicity here)
sig2epsi = (1 - h2) * beta^2 # Variance of the error term epsilon, ensuring desired h2
sigX = sqrt(sig2X) # SD of X
sigepsi = sqrt(sig2epsi)

print(paste("Variance of X:", sig2X))
print(paste("Variance of Epsilon:", sig2epsi))
```

## Generate Single Instance of Data

Simulate vectors X, epsilon (error), and Ynull (outcome under null). Then calculate Yalt.

```{r}
# Simulate predictor X and error term epsi
X = rnorm(nsamp, mean=0, sd= sigX) # Simulating X with specified number of samples, mean and standard deviation
epsi = rnorm(nsamp, mean=0, sd=sigepsi) #simulating our error term epsilon with specified # of samples, mean and SD

# Generate Ynull (independent of X, here just sampling with variance related to beta for comparison)
Ynull = rnorm(nsamp, mean=0, sd=beta) 

# Calculate Yalt = X * beta + epsi
Yalt = X * beta + epsi
```

## Visualize the Data

```{r}
# Visualize Data
plot(X, Yalt, main="Yalt vs X"); grid()
plot(X, Ynull, main="Ynull vs X"); grid()
```

## Test Associations (Single Instance)

Test the association between Ynull and X using linear regression.

```{r}  
summary(lm(Ynull ~ X))
```

Question: What is the p-value of the association between Ynull and X? Is it significant at the α=0.05 level?

Now test the association between Yalt and X.

```{r}
summary(lm(Yalt ~ X))
```

Question: What is the p-value of the association between Yalt and X? Is it significant at the α=0.05 level?

## Calculate the Empirical Distribution of P-values

To understand the distribution of p-values, we need to repeat the simulation many times.

### Convenience Function fastlm

We'll run 10,000 regressions. A simplified function fastlm can speed this up compared to repeated calls to lm().

```{r}
fastlm = function(xx,yy) {
  ## compute betahat (regression coef) and pvalue with Ftest
  ## for now it does not take covariates
  
  df1 = 2 # Degrees of freedom (DFs) for model with predictor
  df0 = 1 # DFs for null model (intercept only)
  ind = !is.na(xx) & !is.na(yy)
  xx = xx[ind]
  yy = yy[ind]
  n = sum(ind)
  
  # Computing means of x and y 
  xbar = mean(xx)
  ybar = mean(yy)
  xx = xx - xbar # Center X
  yy = yy - ybar # Center Y
  
  # Sum of squares for X, y, and the cross-products of X and Y.
  SXX = sum( xx^2 )
  SYY = sum( yy^2 )
  SXY = sum( xx * yy )
  
  # Least squares slope estimate for simple linear regression
  betahat = SXY / SXX
  
  # RSS1 = Residual Sum of Squares for model Y ~ X
  RSS1 = sum( ( yy - xx * betahat )^2 ) 
  # RSS0 = Residual Sum of Squares for model Y ~ 1 (equivalent to total sum of squares of centered Y)
  RSS0 = SYY 
  
  # F-statistic comparing the two models
  fstat = ( ( RSS0 - RSS1 ) / ( df1 - df0 ) )  / ( RSS1 / ( n - df1 ) ) 
  # P-value from F-distribution (one minus the CDF)
  pval = 1 - pf(fstat, df1 = ( df1 - df0 ), df2 = ( n - df1 )) 
  
  res = list(betahat = betahat, pval = pval)
  return(res)
}
```

### Run 10,000 Simulations

Simulate X, Ynull, and Yalt 10,000 times. Store X and the outcomes in matrices.

```{r}
nsim = 10000 # Number of simulations

# Simulate matrices: rows=samples, cols=simulations
Xmat = matrix(rnorm(nsim * nsamp, mean=0, sd= sigX), nsamp, nsim)
epsimat = matrix(rnorm(nsim * nsamp, mean=0, sd=sigepsi), nsamp, nsim)

# Generate Y matrices based on the formulas
Ymat_alt = Xmat * beta + epsimat 
Ymat_null = matrix(rnorm(nsim * nsamp, mean=0, sd=beta), nsamp, nsim) # Null Ys

# Check dimensions
dim(Ymat_null)
dim(Ymat_alt)

# Assign column names for reference
colnames(Ymat_null) = paste0("sim", 1:ncol(Ymat_null))
colnames(Ymat_alt) = colnames(Ymat_null)
```

### Calculate P-values Under the Null

Run regressions for Ymat_null ~ Xmat for each simulation and store the p-values and coefficients.

```{r}
pvec_null = rep(NA, nsim)
bvec_null = rep(NA, nsim)

for(ss in 1:nsim) {
  fit = fastlm(Xmat[,ss], Ymat_null[,ss])
  pvec_null[ss] = fit$pval  
  bvec_null[ss] = fit$betahat
}

summary(pvec_null)
```

```{r}
hist(pvec_null, xlab="p-value", main="Histogram of p-values under Null", breaks=20)
```

Questions:
- The histogram should look approximately uniform. Why?
- How many simulations under the null yield p-value below 0.05? What percentage is that?

```{r}
sum(pvec_null < 0.05)
mean(pvec_null < 0.05)
```

What proportion of simulations do you expect to have p-values < α (for any α between 0 and 1) under the null?
Why does this uniform distribution highlight the need for correction when performing many tests?

# Bonferroni Correction

The simplest correction is the Bonferroni correction.

Method: Divide the desired significance level (e.g., 0.05) by the total number of tests performed (m, here nsim). Use this adjusted value as the new significance threshold.

Goal: Controls the Family-Wise Error Rate (FWER) - the probability of calling one or more significant results across all tests under the null. P(FP>=1) ≤ α.

```{r}
BF_thres = 0.05 / nsim # Bonferroni significance threshold
print(paste("Bonferroni threshold:", BF_thres))

# Number of Bonferroni significant associations under the NULL
print(paste("Found under Null:", sum(pvec_null < BF_thres)))
print(paste("Proportion under Null:", mean(pvec_null < BF_thres)))
```

Questions:
- What is the Bonferroni threshold for significance in this simulation?
- How many false positives (significant results under the null) did we find using this strict threshold?

# Mix of Null and Alternative Hypotheses

Real-world datasets (like GWAS) contain a mixture of true associations (alternative hypothesis) and null results. Let's simulate this.

## Create Mixed Data

Assume a certain proportion (prop_alt) of our simulations represent true associations.

```{r}
prop_alt = 0.20 # Define proportion of alternative Ys in the mixture

# Create a selection vector: 1 if alternative, 0 if null
selectvec = rbinom(nsim, 1, prop_alt) 
names(selectvec) = colnames(Ymat_alt) # Keep track
table(selectvec) # Show counts of null (0) vs alternative (1)

# Create the mixed Y matrix
# If selectvec[i] is 1, use Ymat_alt[,i]; if 0, use Ymat_null[,i]
Ymat_mix = sweep(Ymat_alt, 2, selectvec, FUN='*') + sweep(Ymat_null, 2, 1 - selectvec, FUN='*')
```

## Calculate P-values for Mixed Data

Run regressions for Ymat_mix ~ Xmat.

```{r}
pvec_mix = rep(NA, nsim)
bvec_mix = rep(NA, nsim)

for(ss in 1:nsim) {
  fit = fastlm(Xmat[,ss], Ymat_mix[,ss])
  pvec_mix[ss] = fit$pval  
  bvec_mix[ss] = fit$betahat
}

summary(pvec_mix)
```

```{r}
hist(pvec_mix, xlab="p-value", main="Histogram of p-values under mixture", breaks=20)
```

## Interpreting Mixed P-values

The histogram shows a spike near zero (from true alternatives) superimposed on a uniform background (from true nulls).

```{r}
thres = 0.05
m_signif = sum(pvec_mix < thres)      # Observed number of significant associations
m_expected_null = thres * sum(selectvec == 0) # Expected number of FPs from nulls
m_expected_all_null = thres * nsim # Expected if *all* were null

print(paste("Observed significant (p<0.05):", m_signif))
print(paste("Expected false positives if all were null:", m_expected_all_null))
print(paste("Expected false positives from the actual nulls:", round(m_expected_null)))
```

We observed many more significant results than expected if all tests were null. How can we estimate the proportion of false discoveries among those we call significant?

## False Discovery Rate (FDR) Estimate (using known truth)

Since this is a simulation, we know which tests were truly null (selectvec == 0). We can calculate the actual FDR for a given p-value threshold.

```{r}
# FDR = (Number of significant tests that are truly null) / (Total number of significant tests)
FP = sum(pvec_mix < thres & selectvec == 0) # False Positives
TP = sum(pvec_mix < thres & selectvec == 1) # True Positives
S = sum(pvec_mix < thres) # Total Significant (S = FP + TP)

FDR_sim = FP / S 
print(paste("Simulated FDR at p <", thres, ":", round(FDR_sim, 4)))
```

This means that if we used p < 0.05 as our criterion, about round(FDR_sim*100, 1)% of our significant findings would actually be false positives.

Questions:
- What is the proportion of false discoveries (FDR) if we use a significance level of 0.01?

```{r}
thres_01 = 0.01
FP_01 = sum(pvec_mix < thres_01 & selectvec == 0)
S_01 = sum(pvec_mix < thres_01)
FDR_sim_01 = FP_01 / S_01
print(paste("Simulated FDR at p < 0.01:", round(FDR_sim_01, 4)))
```

- What is the proportion of false discoveries if we use the Bonferroni threshold?

```{r}
FP_bf = sum(pvec_mix < BF_thres & selectvec == 0)
S_bf = sum(pvec_mix < BF_thres)
FDR_sim_bf = ifelse(S_bf > 0, FP_bf / S_bf, 0) # Avoid division by zero
print(paste("Simulated FDR at Bonferroni p <", BF_thres, ":", round(FDR_sim_bf, 4)))
print(paste("Number significant at Bonferroni:", S_bf))
```

- What is the proportion of missed signals (False Negative Rate among true alternatives) using the Bonferroni threshold?

```{r}
FN_bf = sum(pvec_mix >= BF_thres & selectvec == 1) # False Negatives
Total_Alt = sum(selectvec == 1) # Total true alternatives
FNR_sim_bf = FN_bf / Total_Alt
print(paste("Simulated FNR at Bonferroni p <", BF_thres, ":", round(FNR_sim_bf, 4)))
```

# Common Approaches to Control Type I Errors

Let m be the total number of tests. Consider the following table:

|| Called Not Significant | Called Significant | Total
|---|---------------|---------------------|-----
| Null True | TN | FP | $m_0$
| Alternative True | FN | TP | $m_1$
| Total | $m$ - S | S | $m$

(m total # tests, $m_0$ # null tests, FP false positives, TP true positives, TN true negatives, FN false negatives, S total significant)

- FWER (Family-Wise Error Rate): P(FP ≥ 1). Probability of at least one false positive. Controlled by Bonferroni. Often too strict.
- FDR (False Discovery Rate): E[FP / S | S > 0]. Expected proportion of false positives among all significant results. A more liberal approach, often preferred when many tests are performed and finding true positives is important.

## Table of Simulation Results

Let's construct the confusion table for our mixed simulation using alpha = 0.05.

```{r}
count_table = t(table(Significant = pvec_mix < 0.05, True_Status = selectvec))
# Reorder and rename for standard confusion matrix format
# Rows: True Status (0=Null, 1=Alt); Columns: Called Status (FALSE=Not Sig, TRUE=Sig)
conf_matrix = count_table[, c("FALSE", "TRUE")] 
colnames(conf_matrix) = c("Called Not Significant", "Called Significant")
rownames(conf_matrix) = c("Null True", "Alt True")

knitr::kable(conf_matrix, caption="Confusion Matrix at p < 0.05")

# Extracting values
TN = conf_matrix["Null True", "Called Not Significant"]
FP = conf_matrix["Null True", "Called Significant"]
FN = conf_matrix["Alt True", "Called Not Significant"]
TP = conf_matrix["Alt True", "Called Significant"]
print(paste("TN:", TN, " FP:", FP, " FN:", FN, " TP:", TP))
```

# Use qvalue Package to Estimate FDR

In real analysis, we don't know selectvec. The qvalue package estimates FDR and related quantities directly from the p-value distribution.

q-value: The minimum FDR at which a test would be called significant. If you call all tests with q ≤ x significant, you expect the FDR to be approximately x.

```{r}
# Install qvalue package if you haven't already
# if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("qvalue")
```

```{r}
library(qvalue)
```

```{r}
# Calculate q-values for the mixed p-values
qres_mix = qvalue(pvec_mix)
qvec_mix = qres_mix$qvalues

# Also calculate for the null p-values for comparison
qres_null = qvalue(pvec_null)
qvec_null = qres_null$qvalues

summary(qvec_mix)
```

## Visualize Q-values

Let's see if small q-values correspond to true associations (selectvec = 1).

```{r}
boxplot(qvec_mix ~ selectvec, main='Q-value by True Status', xlab="True Status (0=Null, 1=Alt)", ylab="Q-value")
grid()
```

Plot sorted q-values, colored by true status.

```{r}
ind = order(qvec_mix, decreasing=FALSE)
# Using selectvec*2 + 1 maps 0->1 (black) and 1->3 (green) for plotting colors
plot(sort(qvec_mix), col=selectvec[ind]*2 + 1, 
     pch=16, #selectvec[ind] + 1, 
     # lwd=selectvec[ind]*2 + 1,
     main="Sorted Q-values by True Status",
     xlab="Tests ordered by Q-value", ylab="Q-value")
legend("bottomright", legend=c("True Null", "True Alternative"), pch=16, col=c(1, 3))
```

Questions:
- Do small q-values tend to correspond to true alternatives (green points)?
- Interpret the sorted q-value plot: What does the pattern show?

Compare p-value and q-value distributions by true status.

```{r}
par(mfrow=c(1,2))
boxplot(pvec_mix ~ selectvec, main='P-value by True Status', xlab="True Status (0=Null, 1=Alt)", ylab="P-value"); grid()
boxplot(qvec_mix ~ selectvec, main='Q-value by True Status', xlab="True Status (0=Null, 1=Alt)", ylab="Q-value"); grid()
par(mfrow=c(1,1))
```

Question: Interpret these boxplots. How do q-values differ from p-values, especially for the null tests?

## How Do Q-values and P-values Relate?

```{r}
par(mfrow=c(1,2))
plot(pvec_null, qvec_null, main='Q-value vs P-value (All Null)', xlab="P-value", ylab="Q-value"); grid(); abline(0,1,col='red',lty=2)
plot(pvec_mix, qvec_mix, main='Q-value vs P-value (Mixture)', xlab="P-value", ylab="Q-value"); grid(); abline(0,1,col='red',lty=2)
par(mfrow=c(1,1))
```

Q-values are monotonically increasing functions of p-values. Note that q-values are often ≥ p-values. The q-value can be interpreted as the FDR associated with calling that specific test significant.

Questions:
- What is the smallest q-value when all simulations are from the null? Why?
- What is the smallest q-value when simulations are from the mixture?

## Estimating π₀ (Proportion of True Nulls)

The qvalue package also estimates π₀, the overall proportion of features (tests) that are truly null. 

π₁ = 1 - π₀ is the proportion of true alternative hypotheses.

```{r}
# Estimated pi0 when all tests were truly null
print(paste("Estimated pi0 (all null):", round(qres_null$pi0, 4)))

# Estimated pi0 for the mixed data
print(paste("Estimated pi0 (mixture):", round(qres_mix$pi0, 4)))

# True proportion used in simulation
print(paste("True pi0 used:", 1 - prop_alt))
```

Question: How well did the qvalue package estimate the true proportion of null hypotheses (pi0) in our mixed simulation?

# References

Storey, John D., and Robert Tibshirani. 2003. "Statistical Significance for Genomewide Studies." Proceedings of the National Academy of Sciences 100 (16): 9440–45.
